---
title: "BOO 2025 - Example Analysis"
subtitle: "Script 7: TXG-MAPr import - Answers"
date: "`r Sys.Date()`" 
author: 
  Bärbel Smit
output:
  html_document:
    code_download: true
    theme: united
    lowlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

> In this script, you will export your log2FC in a format suitable for upload to the PHH TXG-MAPr.

***

# Setup

## Clean

As before, we perform several setup steps at the start of the script to ensure our work is reproducible and clear. 

**Download the R markdown template, clean your environment, set variables, and load in `tidyverse`:**

* `root_dir` - project folder
* `low_store` - location of DEG results for the low toxicity compound
* `high_store` - location of DEG results for the high toxicity compound

```{r clean}
# Clean environment
rm(list=ls())

# set variables 
root_dir <- "C:/Users/Bärbel Smit/Universiteit Leiden/BOO 2025 - BOO CDS Giulia team - BOO CDS Giulia team/Students/Bärbel/project/barbelsmit_BOO/"


low_store <- "DEG/DEGres_low.Rdata"
high_store <- "DEG/DEGres_high.Rdata"

txg_store <- "TXG_upload"

# Load packages
library(tidyverse)
```

***

## Load data

**Load the DEG results R environment:**

```{r load-data, warning=FALSE, message=FALSE}
load(paste0(root_dir, high_store))
load(paste0(root_dir, low_store))
```

***

# Upload

## File format

Upload files to the TXG-MAPr need the following variables:

* `experiment` defines the experiment (mean_ID). You will be able to look at map activation for each experiment separately.
* `gene_id` defines the gene ID (gene_symbol)
* `time` defines the time of exposure (24 hours for these samples)
* `conc` defines the concentration (dose level)
* `log2fc` specifies the log2 fold change
* `pvalue` defines the p-value
* `padj` defines the adjusted p-value

**Make the data frame and save it as a text file:**

Making text files to calculate EG scores

```{r}
txg_high <- res_high %>% 
  separate(mean_ID, sep="_", into=c("Compound", "conc", NA), remove=F) %>% 
  mutate(time = 24) %>% 
  select(
    experiment = Compound,
    gene_id = gene_symbol,
    time, conc, log2fc = log2FoldChange, pvalue, padj)

write_tsv(txg_high, path=paste0(root_dir, txg_store, "_high.txt"))

txg_low <- res_low %>% 
  separate(mean_ID, sep="_", into=c("Compound", "conc", NA), remove=F) %>% 
  mutate(time = 24) %>% 
  select(
    experiment = Compound,
    gene_id = gene_symbol,
    time, conc, log2fc = log2FoldChange, pvalue, padj)

write_tsv(txg_low, path=paste0(root_dir, txg_store, "_low.txt"))
```

These can then be imported into the PHH TXG-MAPr and explored. 

***

# Session Info

```{r session-info}
library(sessioninfo)

session_info()
```

***

